---
title: "Sugar Forecasting"
Author: Nathan Garn
date: "2023-04-15"
output:
  html_document:
    highlight: breezedark
    number_sections: yes
    toc: yes
    fig_width: 15
    fig_height: 10
  '': default
---

# Packages
```{r packages}
library(quantmod)
library(ggplot2)
library(forecast)
library(astsa)
library(tseries)
library(tidyverse)
library(caret)
library(bigrquery)
library(prophet)
library(lubridate)
library(openxlsx)
library(dplyr)
library(readr)
library(feasts)
library(tsibble)
library(Metrics)
library(rugarch)
library(rmgarch)


```

# Data Load
```{r}

sugar <- read.csv("sugar-prices-historical-chart-data.csv",stringsAsFactors = FALSE)%>%
mutate(date = mdy(date))%>% 
mutate(month = format_ISO8601(date, precision = "ym"))

```

# Exploring Sugar
```{r}

# Looking at the price of sugar over the last 30 years
ggplot(sugar, aes(date, price)) +
  geom_line() +
  labs(title = "Sugar Prices") + xlab("year") + ylab("prices")
  
#Summary of sugar
summary(sugar)
#Significant fluctuations in price.
  
#Sugar monthly average

sugar_m <- sugar %>% group_by(month) %>% mutate(mon_avg = mean(price))%>%
select(month, mon_avg)

#Drop Duplicate rows Sugar

sugar_m <- sugar_m[!duplicated(sugar_m),]

#Creating Time series data for Sugar

sugar_ts <- ts(sugar_m$mon_avg,start=c(1993),frequency=12)

#Plotting

chartSeries(sugar_ts)


#Looking at trends

autoplot(decompose((sugar_ts)),main="") 
#Price spikes occur between 2010-2012 and again in the 2020's

#Sugar Looking for daily or weekly trends

sugar_r <- sugar %>%
filter(date >= as.Date('2022-11-01') & date <= as.Date('2023-01-31'))


ggplot(sugar_r, aes(date, price)) +
  geom_line() +
  labs(title = "Sugar Prices Daily") + xlab("time") + ylab("prices")

# There does not appear to be trends by day of the week. 


#Sugar Prices Controlling for inflation

#write.csv(sugar_m, file="sugar_m.csv",row.names = FALSE)

sugar_CPI <- read.csv("Sugar_cpi.csv")

sugar_adj <- left_join(sugar_m, sugar_CPI, by = c("month" = "date"))

sugar_adj$adj_price <- sugar_adj$mon_avg / sugar_adj$Value

sugar_ts_adj <- ts(sugar_adj$adj_price, start = c(1993), frequency = 12)


ggplot(sugar, aes(date, price)) +
  geom_line() +
  labs(title = "Sugar Prices") + xlab("year") + ylab("prices")

 chartSeries(sugar_ts)

ggplot(data = sugar_adj, aes(x = month, y = adj_price, group = 1)) +
  geom_line() +
  labs(x = "Month", y = "Adjusted Sugar Price", title = "Sugar Prices Adjusted for Inflation")

ggplot(sugar_adj, aes(x = month)) +
  geom_line(aes(y = Value, color = "Inflation Rate", group = 1)) +
  geom_line(aes(y = mon_avg, color = "Sugar Price", group = 1)) +
  scale_color_manual(values = c("blue", "red")) +
  xlab("Month") +
  ylab("Value") +
  ggtitle("Inflation Rate and Sugar Prices over Time")



sugar_adj$norm_value <- scale(sugar_adj$Value)
sugar_adj$norm_mon_avg <- scale(sugar_adj$mon_avg)

# Calculate the adjusted sugar price by dividing the monthly average sugar price by the CPI value


# Plot the normalized values for 'Value' and 'mon_avg' and the adjusted sugar price as lines

#changing month to a date
sugar_adj$month <- as.Date(paste0(sugar_adj$month, "-01"))

ggplot(sugar_adj, aes(x = month)) +
  geom_line(aes(y = norm_value, color = "CPI", group = 1)) +
  geom_line(aes(y = norm_mon_avg, color = "Sugar Price", group = 1)) +
  scale_color_manual(values = c("darkblue", "orange")) +
  xlab("Date") +
  ylab("Normalized Values") +
  ggtitle("Inflation Rate vs. Price - Sugar") +
  scale_x_date(date_labels = "%Y")

#Seasonal Plots
#Format back to date in aggregated month column
sugar_m %>%
  mutate(month = ym(month)) -> Sugar_mon

#Convert to tibble
Sugar_mon <- as_tibble(Sugar_mon)

#convert data frame into time series tsibble object
Sugar_mon %>% as_tsibble(index = month) -> Sugar_mon_ts

#Format data for Sugar seasonal plots
Sugar_mon_ts %>%
  mutate(Month = tsibble::yearmonth(month)) %>%
  as_tsibble(index = Month) %>%
  dplyr::select(Month,mon_avg) -> Sugar_sea_ts

#Sugar seasonal plot by month
autoplot(Sugar_sea_ts, mon_avg) +
  ylab("monthly Sugar prices") +
  xlab("")

#Different view Sugar seasonal plot
Sugar_sea_ts %>% gg_season(mon_avg, labels = "both") + 
  ylab("Monthly Sugar prices")

#Sugar seasonal subseries plot 
Sugar_sea_ts %>% gg_subseries(mon_avg) +
  labs(y = "Sugar prices", title = "Seasonal subseries plot: Sugar prices")
```

# Arima Sugar
```{r}
##Stationary Test
adf.test(sugar_ts, alternative = "stationary")

## After first-order differencing
adf.test(diff(sugar_ts), alternative ="stationary")
#This decreases our p-value and creates significance
#Data is now stationary making the value of (d)=1

#Custom ARIMA Model

#Correlation Plot and Tuning selection. This plot looks at price correlation with itself in prior time periods. The parameter selection controls for that. 
#ACF (q)
acf(diff(sugar_ts),main='')
#q=2

#Looking at impact of price prior periods have on consecutive time periods
#PACF (p)
pacf(diff(sugar_ts),main='')
#p=2. There are 2 partial auto corelation values

# ARIMA Custom

sugar_fit<- Arima(sugar_ts, order=c(2,1,2))

BIC(sugar_fit)
# Calculate the AICc value

loglik_sugar_fit <- logLik(sugar_fit)

loglik_sugar_fit <- logLik(sugar_fit) - max(loglik_sugar_fit)

n <- length(sugar_ts)
k <- length(coef(sugar_fit))
sugar_fit_AICc <- AIC(sugar_fit) + 2*k*(k+1)/(n-k-1) - 2*loglik_sugar_fit

sugar_fit_AICc

sugar_fit

# ARIMA Alternate Custom

sugar_fit2<- Arima(sugar_ts, order=c(2,1,3))
sugar_fit2

forecast::accuracy(sugar_fit2)

#Check residuals

checkresiduals(sugar_fit2)

#Auto-fit Arima

auto_sugar<- auto.arima(sugar_ts)
auto_sugar

##Forecast Plot

##Forecast Custom

autoplot(forecast::forecast(sugar_fit, h=12, level=c(80,95)))

##Forecast Custom 2

autoplot(forecast::forecast(sugar_fit2, h=12, level=c(80,95)))

##Forecast Auto

autoplot(forecast::forecast(auto_sugar, h=12, level=c(80,95)))


#ARIMA using more recent data

sugar_r2 <- sugar %>%
filter(date >= as.Date('2019-01-01') & date <= as.Date('2023-01-31'))

sugar_r2 <- sugar_r2 %>% group_by(month) %>% mutate(mon_avg = mean(price))%>%
select(month, mon_avg)

#Drop Duplicate rows Sugar

sugar_r2 <- sugar_r2[!duplicated(sugar_r2),]

sugar_tsr <- ts(sugar_r2$mon_avg,start=c(2019),frequency=12)

##Stationary Test
adf.test(sugar_tsr, alternative = "stationary")

## After first-order differencing
adf.test(diff(sugar_tsr), alternative ="stationary")
#This decreases our p-value and creates significance
#Data is now stationary making the value of (d)=2

#Custom Arima Model

#Correlation Plot and Tuning selection
#ACF (q)
acf(diff(sugar_tsr),main='')
#q=0

#Looking at impact of price prior periods have on consecutive time periods
#PACF (p)
pacf(diff(sugar_tsr),main='')
#p=0. There are 2 partial auto corelation values

# ARIMA Custom

sugar_fitR<- Arima(sugar_tsr, order=c(0,2,0))
sugar_fitR

forecast::accuracy(sugar_fitR)

#Check residuals

checkresiduals(sugar_fitR)

#Auto-fit Arima

auto_sugarR<- auto.arima(sugar_tsr)
auto_sugarR

##Forecast Plot

autoplot(forecast::forecast(auto_sugarR, h=12, level=c(80,95)))

##Forecast Custom

autoplot(forecast::forecast(sugar_fitR, h=12, level=c(80,95)))

#Printing Predictions

sugar_predictions <- forecast::forecast(auto_sugarR,h=12)

print(sugar_predictions$mean)
```
# Sugar GARCH
```{r}

# Model Creation

garch_model <- ugarchspec(variance.model = list(model = "sGARCH", 
                                                garchOrder = c(1,1)), 
                                                mean.model = list(armaOrder =
                                                c(0,0), include.mean = TRUE), distribution.model = "std")
# Fitting Model to Data

sugar_garch <- ugarchfit(spec = garch_model, data = sugar_ts)

sugar_vol <-ts(sugar_garch@fit$sigma^2,start=c(1993),frequency=12)


print(sugar_garch)

#plot(sugar_garch, which = 1)

coef(sugar_garch)

# Forecasting

horizon <- 3

sugar_forecast_garch <- ugarchforecast(sugar_garch, n.ahead = horizon)

forecast_mean_sugar <- as.numeric(sugar_forecast_garch@forecast$seriesFor)
actual_values_sugar <- as.numeric(window(sugar_vol, start = c(1993, 1)))


plot(sugar_forecast_garch, n.plot = horizon, n.col = 1, plot.type = "single", 
     main = "GARCH Forecast for Sugar Prices", ylab = "Price", xlab = "Time") #%>%
lines(sugar_ts[(length(sugar_ts)-horizon+1):length(sugar_ts)], col = "blue")


#GARCH Take 2
#Use ARIMA values from acf, pacf. Did this in Sugar_fit
sugar_fit
sugar_fit2
sugar_fitR
auto_sugar
auto_sugarR

#Based on significance. Let's try auto_arimas

garch_model2 <- ugarchspec(variance.model = list(model = "sGARCH", 
                                                garchOrder = c(1,1)), 
                                                mean.model = list(armaOrder =
                                                c(1,0), include.mean = TRUE), distribution.model = "std")

sugar_garch2 <- ugarchfit(spec = garch_model, data = sugar_ts)
sugar_garch2

#Time Series based on volatility or Variance based on a standard Garch [1,1] model

sugar_vol <-ts(sugar_garch2@fit$sigma^2,start=c(1993),frequency=12)

plot(sugar_vol,xlab="",ylab="",main="Sugar_Volatility (GARCH[1,1])")

#Exponential GARCH (does not work quite as well)
Egarch_model <- ugarchspec(variance.model = list(model = "eGARCH", 
                                                garchOrder = c(1,1)), 
                                                mean.model = list(armaOrder =
                                                c(1,0), include.mean = TRUE), distribution.model = "std")

sugar_egarch2 <- ugarchfit(spec = garch_model, data = sugar_ts)
sugar_egarch2
coef(sugar_egarch2)

sugar_forecast_garche <- ugarchforecast(sugar_egarch2, n.ahead = horizon)

forecast_mean_sugare <- as.numeric(sugar_forecast_garche@forecast$seriesFor)
actual_values_sugare <- as.numeric(window(sugar_vol, start = c(1993, 1)))

#Time Series based on volatility or Variance based on a standard Garch [1,1] model

esugar_vol <-ts(sugar_egarch2@fit$sigma^2,start=c(1993),frequency=12)

#plot(esugar_vol,xlab="",ylab="",main="Sugar_Volatility (eGARCH[1,1])")

cor(sugar_vol, esugar_vol)

ts.plot(sugar_vol,esugar_vol,col=c("green","red"),xlab="")
legend("topright",legend=c("Standard","Exponential"),col=c("green","red"),lty=c(1,1))
#No difference shown

#GARCH 3

names(sugar_garch2@model)
names(sugar_garch2@fit)
#Variance
sugar_garch_var <- sugar_garch2@fit$var
#Residuals
sugar_garch_res <- (sugar_garch2@fit$residuals)^2

#Plotting residuals and conditional variances
plot(sugar_garch_res, type = "l")
lines(sugar_garch_var, col = "green")

sugar_forecast_garch2 <- ugarchforecast(sugar_garch2, n.ahead = 12)
sugar_forecast_garch2

sugar_forecast_values <- as.numeric(sugar_forecast_garch2@forecast$series)
print(sugar_forecast_values)
sugar_forecast_index <- sugar_forecast_garch2@forecast$seriesFor
sugar_forecast <- data.frame(time = sugar_forecast_index, forecast = sugar_forecast_values)
print(sugar_forecast)
#write.csv(sugar_forecast, file="sugar_forecast.csv",row.names = FALSE)

sugar_garch2_fitted <- fitted(sugar_garch2)
#print(sugar_garch2_fitted)


#print(sugar_forecast_values)

#summary(sugar_forecast_garch2)

ug_sugar <- sugar_forecast_garch2@forecast$sigmaFor
plot(ug_sugar, type = "l")

sug_var_t <- c(tail(sugar_garch_var,20),rep(NA,10))  # gets the last 20 observations
sug_res_t <- c(tail(sugar_garch_res,20),rep(NA,10))  # gets the last 20 observations
sug_f <- c(rep(NA,20),(ug_sugar)^2)

plot(sug_res_t, type = "l") #Residuals
lines(sug_f, col = "orange") # Predictions 
lines(sug_var_t, col = "green") #Conditional Forecast


#Plot Predictions

sug_mean_forecast <- as.numeric(sugar_forecast_garch2@forecast$seriesFor)

# Get the upper and lower confidence intervals for both 95% and 80%
sug_conf_int_95 <- as.numeric(sugar_forecast_garch2@forecast$upper[, "95%"]) # 95% confidence interval
sug_conf_int_80 <- as.numeric(sugar_forecast_garch2@forecast$upper[, "80%"]) # 80% confidence interval

# Plot the mean forecasted values with the two confidence intervals

#plot(sugar_forecast_garch2, main = "Forecasted coffee Prices (GARCH(1,1))")


```

# Sugar Model Comparison
```{r}

#ARIMA Models
forecast::accuracy(sugar_fit)
forecast::accuracy(sugar_fit2)
forecast::accuracy(auto_sugar)
forecast::accuracy(sugar_fitR)
forecast::accuracy(auto_sugarR)

AIC(sugar_fit2)
BIC(sugar_fit2)

#GARCH Model

actual_values_sugar <- as.numeric(window(sugar_ts))
actual_values_sugar <- head(actual_values_sugar, length(forecast_mean_sugar))

mae <- mean(abs(forecast_mean_sugar - actual_values_sugar))
mse <- mean((forecast_mean_sugar - actual_values_sugar)^2)
rmse <- sqrt(mse)
# Print the results
cat(paste("MAE: ", mae, "\n"))
cat(paste("MSE: ", mse, "\n"))
cat(paste("RMSE: ", rmse, "\n"))

#GARCH Model 2

forecast_mean_sugar2 <- as.numeric(sugar_forecast_garch2@forecast$seriesFor)
actual_values_sugar2 <- as.numeric(window(sugar_ts))
actual_values_sugar2 <- head(actual_values_sugar2, length(forecast_mean_sugar2))


sug_mae <- mean(abs(forecast_mean_sugar2 - actual_values_sugar2))
sug_mse <- mean((forecast_mean_sugar2 - actual_values_sugar2)^2)
sug_rmse <- sqrt(sug_mse)
# Print the results
cat(paste("MAE: ", sug_mae, "\n"))
cat(paste("MSE: ", sug_mse, "\n"))
cat(paste("RMSE: ", sug_rmse, "\n"))


#eGARCH

forecast_mean_sugare <- as.numeric(sugar_forecast_garche@forecast$seriesFor)
actual_values_sugare <- as.numeric(window(sugar_ts))
actual_values_sugare <- head(actual_values_sugar2, length(forecast_mean_sugare))


esug_mae <- mean(abs(forecast_mean_sugare - actual_values_sugare))
esug_mse <- mean((forecast_mean_sugare - actual_values_sugare)^2)
esug_rmse <- sqrt(sug_mse)
# Print the results
cat(paste("MAE: ", esug_mae, "\n"))
cat(paste("MSE: ", esug_mse, "\n"))
cat(paste("RMSE: ", esug_rmse, "\n"))
